---
title:  "A review of statistical computing with TMB"
author: "Andrea Havron<br>NOAA Fisheries, OST"
output: html_document
---

# ML Inference

What is the likelihood of getting heads 30 times in 100 coin flips?


1. Specify the model <br><br>
$y ~ \sim Binomial(n, p)$


2. Calculate the likelihood<br><br>
$L(p; n, y) = \frac{n!}{y!(n-y)!}p^y(1-p)^{n-y}$

<br>
$L(p; n = 100, y = 30)$
```{r, eval=TRUE, out.width = '80%', echo = FALSE}
curve(dbinom(30,100,x),0,0.75, ylab = 'L(p)', xlab = 'p')
```


3. Calculate the negative log-likelihood<br><br>

$-\ell(p; n, y) = -[ln\big(\frac{n!}{y!(n-y)!}\big) + yln(p) + (n-y)ln(1-p)]$

<br>

$-ln\big[L(p; n = 100, y = 30)\big]$
```{r, eval=TRUE, out.width = '80%', echo = FALSE}
curve(-log(dbinom(30,100,x)),0,0.75, ylab = 'l(p)', xlab = 'p')
```

4. Calculate the derivative w.r.t. $p$<br><br>
$\frac{d(\ell(p; n, y))}{dp} = \frac{y}{p}- \frac{n-y}{1-p}$

<br>

$-ln\big[L(p; n = 100, y = 30)\big]$

```{r, eval=TRUE, out.width = '80%', echo = FALSE}
curve(-log(dbinom(30,100,x)),0,0.75, ylab = 'l(p)', xlab = 'p')
```

5. Set to 0 and solve for MLE<br>
\begin{align}
0 &= \frac{y}{p}- \frac{n-y}{1-p}\\
E[p] &= \frac{y}{n}\\
E[y] &= np
\end{align}

<br>

$-ln\big[L(p; N = 100, y = 30)\big]$

<br>

```{r, eval=TRUE, out.width = '70%', echo = FALSE}
curve(-log(dbinom(30,100,x)),0,0.75, ylab = 'l(p)', xlab = 'p')
nll <- -dbinom(30,100,.3,TRUE)
segments(0.1, nll, 0.5, nll, col='red', lwd=2)
```

<br>

$\hat{p} = \frac{30}{100} = 0.3$

<br>

6. Approximate Var[p] using the second derivative<br>

\begin{align}
&-\frac{y}{p^2} - \frac{(n-y)}{(1-p)^2}\\
&-\frac{np}{p^2} - \frac{(n-np)}{(1-p)^2}\\
&-\frac{n}{p} - \frac{n}{1-p}\\
l''(p) &= -\frac{n(1-p)}{p}\\
Var[p] &= \frac{p(1-p)}{n}\\
\end{align}

<br>

\begin{align}
Var[p] &\approx -\frac{1}{l''(p)}\\
SE[p] &\approx \sqrt{ \frac{(.3)(.7)}{100}} \approx 0.00458\\
\end{align}

<br>

```{r, eval=TRUE, out.width = '70%', echo = FALSE}
curve(-log(dbinom(30,100,x)),0,0.75, ylab = 'l(p)', xlab = 'p')
nll <- -dbinom(30,100,.3,TRUE)
points(0.3, nll, col = 'red')
confint <- .3 + c(-1,1)*qnorm(.975)*sqrt(.3*.7/100)
segments(confint[1], -log(dbinom(30,100,confint[1])), confint[1], 100, col='red', lwd=2)
segments(confint[2], -log(dbinom(30,100,confint[2])), confint[2], 100, col='red', lwd=2)
text(.3, 20, "95% confint", col='red', cex = 2)
```
